---
title: "FEWS farm model: `r params$label`"
author: "Jones (2022), Environment Systems and Decisions"
date: "`r format(Sys.Date())`"
params:
  site: EP
  label: Equally Probable
  root: ".."
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE, eval = TRUE, fig.align = "center",
  warning = FALSE, message = FALSE
)
```

```{r libraries, include = FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)

# The seed does not affect any number in this report -- nothing here samples.
# It is set because the report is about a stochastic model and a reader is
# entitled to see that the randomness lives in the committed input file rather
# than in this session. See docs/reproduction-notes.md.
set.seed(12345)
```

<!--
This one file replaces eleven.

  FM DML Graphs.Rmd            FM EP Graphs.Rmd
  FM DML MC.Rmd                FM EP MC.Rmd
  FM Traditional DML.Rmd       FM Traditional EP.Rmd
  FM MI DML All Climates.Rmd   FM MI EP All Climates.Rmd
                               FM MI EP 1 Climate.Rmd
  FM Final Outputs.Rmd         FM output combine.Rmd

They were a 2 x 5 grid built by copy-paste, differing in a site name, plus two
files that combined the results. Between them they carried more than thirty
absolute paths to a machine layout that no longer exists -- nine in
`FM Final Outputs.Rmd` alone -- and eleven copies of the same aggregation, so a
correction to one of them reached none of the others.

Every path here is derived from `params$root`, which is how the duplication was
removed rather than merely reduced: a path cannot stay hardcoded once the site is
an argument.

This report READS what stage 1 wrote and does not solve anything. The model lives
in `src/fews_stochopt/`, is exercised by `scripts/run_all.py`, and is the single
source of every number below.
-->

# What this report covers

`r params$label` (`r params$site`) is one of the paper's two climate regimes. Four
scenarios differ only in what the farm knows when it chooses its alternative water
and electricity capacity:

| Scenario | The farm invests knowing |
|---|---|
| Perfect Information | exactly the weather it is about to get |
| Known Climate, Unknown Weather | which climate it is in, not which years |
| Stochastic | the distribution over all climates |
| Expected Value | one deterministic precipitation path |

The differences between their expected profits are the value-of-information
figures the paper reports.

```{r paths}
root <- normalizePath(params$root, mustWork = TRUE)
results <- file.path(root, "results", params$site)

stopifnot(dir.exists(results))

slug <- function(scenario) {
  gsub(" ", "_", gsub(",", "", tolower(scenario)))
}

scenarios <- c(
  "Perfect Information",
  "Known Climate, Unknown Weather",
  "Stochastic",
  "Expected Value"
)

read_part <- function(scenario, part) {
  path <- file.path(results, paste0(slug(scenario), "_", part, ".csv"))
  if (!file.exists(path)) {
    stop(
      "missing ", basename(path), " in ", results, ".\n",
      "Stage 1 has not run for this site. From the repository root:\n",
      "    python scripts/run_all.py",
      call. = FALSE
    )
  }
  read_csv(path, show_col_types = FALSE)
}
```

# Stage 1 output

```{r load}
profit <- lapply(scenarios, read_part, part = "profit")
detail <- lapply(scenarios, read_part, part = "detail")
invest <- lapply(scenarios, read_part, part = "investment")
names(profit) <- names(detail) <- names(invest) <- scenarios

# A silently short read would give a plausible mean and a wrong table, so check
# the shape rather than the absence of an error.
runs <- nrow(profit[["Stochastic"]])
years <- length(unique(detail[["Stochastic"]]$year))
stopifnot(runs > 0, years > 0)
for (s in scenarios) {
  stopifnot(nrow(profit[[s]]) == runs)
  stopifnot(nrow(detail[[s]]) == runs * years)
}

cat(sprintf("%d Monte Carlo runs x %d years, in each of %d scenarios\n",
            runs, years, length(scenarios)))
```

# Profit, by scenario

The whole of Table 5 is mean, standard deviation and a 95% interval for profit and
crop yield. Two details of the original are reproduced rather than corrected,
because the published numbers carry them: the standard error divides by
`sqrt(n - 1)`, and profit is summarised over runs while crop yield is summarised
over run-years, so the two carry different degrees of freedom in the same row.

```{r stats}
summarise_vector <- function(x) {
  n <- length(x)
  sd_x <- sd(x)
  t <- -qt(0.025, df = n - 1)
  se <- sd_x / sqrt(n - 1)
  c(mean = mean(x), sd = sd_x, t = t, se = se, h = t * se)
}

sim_stats <- bind_rows(lapply(scenarios, function(s) {
  p <- summarise_vector(profit[[s]]$profit)
  y <- summarise_vector(detail[[s]]$crop_yield)
  tibble(
    sim = s,
    profit_mean = p[["mean"]], profit_sd = p[["sd"]], profit_t = p[["t"]],
    profit_se = p[["se"]], profit_h = p[["h"]],
    crop_yield_mean = y[["mean"]], crop_yield_sd = y[["sd"]],
    crop_yield_t = y[["t"]], crop_yield_se = y[["se"]], crop_yield_h = y[["h"]]
  )
})) %>%
  mutate(sim = factor(sim, levels = scenarios))

knitr::kable(
  sim_stats %>% select(sim, profit_mean, profit_sd, profit_h,
                       crop_yield_mean, crop_yield_sd),
  digits = 4,
  caption = paste("Table 5 for", params$label)
)
```

```{r profit-plot, fig.width = 8, fig.height = 4.5}
# Alt text: four horizontal 95% confidence intervals for mean profit, one per
# scenario, ordered Perfect Information at the top down to Expected Value at the
# bottom. Perfect Information is highest at about $2.36M for Equally Probable and
# $1.98M for Dry Most Likely; Known Climate sits just below it; Stochastic and
# Expected Value are lower and lie almost exactly on top of each other. The
# intervals are far narrower than the gaps between the first three, and the gap
# between the last two is invisible at this scale -- which is the point the value
# of the stochastic solution makes numerically below.
ggplot(sim_stats, aes(x = profit_mean, y = sim)) +
  geom_errorbarh(aes(xmin = profit_mean - profit_h, xmax = profit_mean + profit_h),
                 height = 0.2) +
  geom_point(size = 2.5, shape = 21, fill = "white") +
  scale_x_continuous(labels = function(x) sprintf("%.2fM", x / 1e6)) +
  labs(
    title = paste("Mean profit by scenario --", params$label),
    subtitle = "point and 95% interval; shape and position carry the meaning, not colour",
    x = "Mean profit ($)", y = NULL
  ) +
  theme_minimal(base_size = 12)
```

The numbers behind that figure are the `profit_mean` and `profit_h` columns of the
table above, so a reader who cannot see the plot is not asked to take it on trust.

# The investment decision

```{r investment}
first_stage <- bind_rows(lapply(scenarios, function(s) {
  inv <- invest[[s]]
  tibble(
    sim = s,
    alt_water_cap_cm = weighted.mean(inv$alt_water_cap, inv$n_runs),
    alt_elc_cap_kW = weighted.mean(inv$alt_elc_cap, inv$n_runs),
    groups = nrow(inv)
  )
})) %>%
  mutate(sim = factor(sim, levels = scenarios))

knitr::kable(first_stage, digits = 6,
             caption = "First-stage capacity, run-weighted across groups")
```

`groups` is how many separate investment decisions the scenario makes: one per run
under perfect information, one per climate when the climate is known, one
otherwise. That column *is* the difference between the four scenarios -- everything
else about the model is identical.

# Value of information

```{r values}
mean_profit <- vapply(scenarios, function(s) mean(profit[[s]]$profit), numeric(1))

values <- tibble(
  Climate_Probability = params$label,
  Value_of_Known_Weather = mean_profit[["Perfect Information"]] -
    mean_profit[["Known Climate, Unknown Weather"]],
  Value_of_Perfect_Information = mean_profit[["Perfect Information"]] -
    mean_profit[["Stochastic"]],
  Value_of_Stochastic_Solution = mean_profit[["Stochastic"]] -
    mean_profit[["Expected Value"]],
  Value_of_Known_Climate = mean_profit[["Known Climate, Unknown Weather"]] -
    mean_profit[["Stochastic"]]
)

# Assert the theory. More information cannot be worth less than none, and EVPI is
# EVKW and EVKC by construction. If either fails the scenarios were mixed up, and
# four plausible numbers would otherwise hide it.
stopifnot(
  mean_profit[["Perfect Information"]] >=
    mean_profit[["Known Climate, Unknown Weather"]],
  mean_profit[["Known Climate, Unknown Weather"]] >= mean_profit[["Stochastic"]],
  abs(values$Value_of_Perfect_Information -
        (values$Value_of_Known_Weather + values$Value_of_Known_Climate)) < 1e-6
)

knitr::kable(values, digits = 4,
             caption = paste("Table 4 for", params$label))
```

`Value_of_Known_Climate` is the column the original pipeline never produced. It is
in the paper -- $98,328.78 and $64,865.98 -- and is the part of the value of
perfect information that comes from learning *which climate you are in*, as
opposed to which years within it.

# Agreement with the Python stage

The package computes this same table, and this report recomputes it from the same
per-run output. Two implementations of arithmetic small enough to look obviously
right is exactly where a difference hides, so the two are compared rather than
assumed to agree.

```{r agreement}
python_table <- read_csv(file.path(root, "results", "solnvalues.csv"),
                         show_col_types = FALSE) %>%
  filter(Climate_Probability == params$label)

stopifnot(nrow(python_table) == 1)

fields <- c("Value_of_Known_Weather", "Value_of_Perfect_Information",
            "Value_of_Stochastic_Solution", "Value_of_Known_Climate")

comparison <- tibble(
  field = fields,
  r = as.numeric(values[fields]),
  python = as.numeric(python_table[fields]),
  difference = r - python
)

# Both sides read the same CSVs and do the same arithmetic in double precision.
# The tolerance is set by floating-point summation order, not by the model, so it
# is tight on purpose: anything larger means the two implementations disagree
# about what they are computing.
worst <- max(abs(comparison$difference))
stopifnot(worst < 1e-6)

knitr::kable(comparison, digits = 8,
             caption = "R report against the Python package")
cat(sprintf("largest disagreement: %.3e dollars\n", worst))

write_csv(values, file.path(results, "report_table.csv"))
```

# Where these numbers come from

```{r provenance}
meta_files <- list.files(results, pattern = "\\.meta\\.json$", full.names = TRUE)
stopifnot(length(meta_files) > 0)

provenance <- bind_rows(lapply(meta_files, function(f) {
  txt <- paste(readLines(f, warn = FALSE), collapse = " ")
  field <- function(name) {
    m <- regmatches(txt, regexpr(paste0('"', name, '"\\s*:\\s*"[^"]*"'), txt))
    if (length(m) == 0) NA_character_ else sub('.*:\\s*"([^"]*)"', "\\1", m)
  }
  tibble(scenario = field("scenario"), gurobi = field("gurobi"),
         config = field("config_digest"), source = field("source_digest"))
}))

knitr::kable(provenance, caption = "Solver and code version behind each scenario")
```

Every file read by this report was written by `scripts/run_all.py`, and each
carries the digest of the configuration and the source that produced it. Re-running
that script with any of them changed re-solves rather than reusing the result.
