---
title: "Weather scenario generation: `r params$label`"
author: "Jones (2022), Environment Systems and Decisions"
date: "`r format(Sys.Date())`"
params:
  site: EP
  label: Equally Probable
  root: ".."
  seed: 12345
  iters: 1000
  year_length: 25
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center",
                      warning = FALSE, message = FALSE)
```

```{r libraries, include = FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
```

# What this generates, and what it does not

This is the step that produces the precipitation scenarios the optimisation reads.
It **writes to `results/regenerated/` and never touches
`stage1-python/precips_c0_*.csv`.** Those files are the input of record: they are
what the published run used, and every number in the paper follows from them. An
input that a later stage can overwrite is not an input.

Two changes from the version this replaces, both forced:

- **It reads `reference/reconstructed/trans_matrix_<site>.csv`, not
  `~/Coding/Data/Farm_Model/trans_matrix.csv`.** The latter exists nowhere -- it
  was already gone when this repository was split out, and every R file in the
  original *reads* it while none creates it. The reconstruction is an estimate
  built from the committed draws by `scripts/reconstruct_markov.py`; see
  `reference/reconstructed/README.md` for what it can and cannot recover.
- **The `markovchain` and `diagram` packages are gone.** A walk over a transition
  matrix is four lines of base R, and dropping two dependencies from a step that
  is now reproducible is worth more than the convenience.

**The output of this document is a new sample, not the committed one.** The
original run's RNG state was never recorded, so its exact draws cannot be
recovered; what can be checked is that a fresh sample has the same distribution,
which the last section does.

```{r paths}
root <- normalizePath(params$root, mustWork = TRUE)
set.seed(params$seed)

matrix_path <- file.path(root, "reference", "reconstructed",
                         sprintf("trans_matrix_%s.csv", params$site))
if (!file.exists(matrix_path)) {
  stop("missing ", matrix_path, ".\nFrom the repository root:\n",
       "    python scripts/reconstruct_markov.py", call. = FALSE)
}

tm <- read.csv(matrix_path, row.names = 1, check.names = FALSE) %>% as.matrix()
stopifnot(nrow(tm) == 20, ncol(tm) == 20)
stopifnot(all(abs(rowSums(tm) - 1) < 1e-9))

cat(sprintf("seed %d, %d iterations of %d years\n",
            params$seed, params$iters, params$year_length))
```

# The four climates

The 20x20 matrix is block diagonal: a run never changes climate, so only the four
5x5 diagonal blocks are ever used. This is what the original did with
`tma[1:5,1:5]` and its three siblings, written as a loop instead of four times.

```{r blocks}
state_names <- c("w1", "w2", "w3", "w4", "w5")
n_climates <- 4

blocks <- lapply(seq_len(n_climates), function(k) {
  idx <- ((k - 1) * 5 + 1):(k * 5)
  block <- tm[idx, idx, drop = FALSE]
  dimnames(block) <- list(state_names, state_names)
  block
})

off_diagonal_mass <- sum(tm) - sum(vapply(blocks, sum, numeric(1)))
stopifnot(abs(off_diagonal_mass) < 1e-9)

knitr::kable(round(blocks[[1]], 4),
             caption = "Climate 1 transition probabilities")
```

Each block's rows are identical to within sampling error, which is the finding
recorded in `reference/reconstructed/README.md`: the climate fixes a weather
distribution and each year is an independent draw from it.

# Climate probabilities

The two sites differ **only** in how often each climate occurs. The transition
matrices are the same; the mixture is not.

```{r probabilities}
# Read from config.yaml so this document holds no second copy of them.
config_lines <- readLines(file.path(root, "config.yaml"), warn = FALSE)
site_line <- grep(sprintf("^  %s:\\s*$", params$site), config_lines)
stopifnot(length(site_line) == 1)
prob_line <- grep("climate_probabilities:", config_lines)
prob_line <- prob_line[prob_line > site_line][1]
stopifnot(!is.na(prob_line))

probs <- as.numeric(strsplit(
  gsub(".*\\[|\\].*", "", config_lines[prob_line]), ",")[[1]])
stopifnot(length(probs) == n_climates, abs(sum(probs) - 1) < 1e-12)

knitr::kable(
  data.frame(climate = seq_len(n_climates), probability = probs,
             runs = round(probs * params$iters * n_climates)),
  caption = paste("Climate mixture for", params$label)
)
```

# Weather states in centimetres

```{r states}
cm_in <- 2.54
rainfall_days_perc <- 0.70
inches <- c(w1 = 5, w2 = 15, w3 = 30, w4 = 45, w5 = 60)

# Rounded to two decimals, which is not cosmetic. The original converted states
# to numbers through `gsub('w2', as.character(w2p), ...)`, and as.character()
# prints seven significant digits -- so the committed file holds exactly 26.67
# while the unrounded product is 26.669999999999998. Comparing the two by value
# then finds no matches at all, and the comparison below silently becomes NA
# rather than large. Reproducing the round-trip keeps both sides on the same grid.
depths <- round(inches * cm_in * rainfall_days_perc, 2)

knitr::kable(data.frame(state = names(depths), inches = inches, cm = depths),
             row.names = FALSE, caption = "The five weather states")
```

# Drawing the sequences

```{r draw}
# A walk over a transition matrix. The original called markovchain::rmarkovchain;
# this is the same thing without the dependency, and it stays correct if the
# matrix ever stops being row-homogeneous.
walk <- function(block, n, start = NULL) {
  if (is.null(start)) {
    # Start from the chain's stationary distribution rather than a fixed state,
    # which is what a memoryless chain implies and what avoids a burn-in.
    start <- sample.int(5, 1, prob = colSums(block) / sum(block))
  }
  out <- integer(n)
  state <- start
  for (i in seq_len(n)) {
    state <- sample.int(5, 1, prob = block[state, ])
    out[i] <- state
  }
  out
}

draw_climate <- function(k, n_runs) {
  block <- blocks[[k]]
  seqs <- t(vapply(seq_len(n_runs),
                   function(i) walk(block, params$year_length),
                   integer(params$year_length)))
  matrix(depths[seqs], nrow = n_runs)
}

runs_per_climate <- round(probs * params$iters * n_climates)
stopifnot(sum(runs_per_climate) == params$iters * n_climates)

# Concatenate the climates in order, then renumber 1..4000 -- which is exactly
# how the original built precips_c0, and why the run index encodes the climate.
sample_blocks <- lapply(seq_len(n_climates),
                        function(k) draw_climate(k, runs_per_climate[k]))
precips <- do.call(rbind, sample_blocks)
stopifnot(nrow(precips) == params$iters * n_climates)
stopifnot(ncol(precips) == params$year_length)

regenerated <- as_tibble(precips, .name_repair = ~ as.character(seq_along(.))) %>%
  mutate(run = row_number()) %>%
  pivot_longer(-run, names_to = "year", values_to = "precip") %>%
  mutate(year = as.integer(year)) %>%
  arrange(run, year)

cat(sprintf("drew %d runs x %d years = %d rows\n",
            max(regenerated$run), params$year_length, nrow(regenerated)))
```

# Does the fresh sample match the committed one?

The original's RNG state was never recorded, so these draws are not the committed
ones and cannot be. What *is* checkable is that they come from the same
distribution. If the reconstruction of the transition matrix were wrong, this is
where it would show.

```{r compare, fig.width = 8, fig.height = 4}
committed <- read_csv(
  file.path(root, "stage1-python", sprintf("precips_c0_%s.csv", params$site)),
  show_col_types = FALSE
) %>%
  select(run, year, precip) %>%
  mutate(run = as.integer(run), year = as.integer(year))

comparison <- bind_rows(
  committed %>% mutate(source = "committed"),
  regenerated %>% mutate(source = "regenerated")
)

shares <- comparison %>%
  count(source, precip) %>%
  group_by(source) %>%
  mutate(share = n / sum(n)) %>%
  ungroup() %>%
  select(-n) %>%
  pivot_wider(names_from = source, values_from = share) %>%
  arrange(precip) %>%
  mutate(difference = regenerated - committed)

knitr::kable(shares, digits = 5,
             caption = "Share of run-years in each weather state")

# Both sides must cover all five states. If a value appeared on one side only,
# pivot_wider would give it an NA and `max` would return NA -- which stopifnot
# rejects, but with a message about the comparison rather than about the states.
# Say what is actually wrong instead.
stopifnot(nrow(shares) == 5)
if (anyNA(shares$difference)) {
  stop("a weather state appears on only one side of the comparison:\n",
       paste(utils::capture.output(print(shares)), collapse = "\n"), call. = FALSE)
}

worst <- max(abs(shares$difference))
cat(sprintf("largest difference in state share: %.4f\n", worst))
cat(sprintf("mean precipitation  committed %.4f cm   regenerated %.4f cm\n",
            mean(committed$precip), mean(regenerated$precip)))

# Sampling error on a share from 100,000 draws is about 0.0016, so a difference
# of more than 0.01 is not sampling -- it is the matrix.
stopifnot(is.finite(worst), worst < 0.01)
```

```{r compare-plot, fig.width = 8, fig.height = 4}
# Alt text: paired bars showing the share of run-years falling in each of the five
# weather states, for the committed draws and for this fresh sample. The pairs are
# visually indistinguishable at every state; the table above gives the numbers,
# the largest difference being under one percentage point.
ggplot(comparison, aes(x = factor(round(precip, 2)), fill = source)) +
  geom_bar(aes(y = after_stat(prop), group = source),
           position = position_dodge(0.8), width = 0.7) +
  labs(
    title = paste("Weather state frequencies --", params$label),
    subtitle = "committed input against a fresh draw from the reconstructed matrix",
    x = "Precipitation (cm)", y = "Share of run-years", fill = NULL
  ) +
  theme_minimal(base_size = 12)
```

```{r write}
out_dir <- file.path(root, "results", "regenerated")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
out <- file.path(out_dir, sprintf("precips_c0_%s_regenerated.csv", params$site))

# The name says regenerated, and the directory is not the input directory. Both
# are deliberate: overwriting stage1-python/ would destroy the only record of what
# the published run actually read.
write_csv(regenerated, out)
cat("wrote", out, "\n")
```
